---
title: "RNA-seq downstream analysis"
output:
  html_document:
    toc: true
    toc_float: true
params:
  config_path: "config/example_config.R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source(file.path("R", "helpers.R"))
config <- load_analysis_config(params$config_path)
paths <- prepare_output_directories(config$output_dir)
```

# Purpose

This report performs a declared bulk RNA-seq downstream analysis from integer gene counts and sample metadata. Comparisons come only from the supplied contrast table. The data-status label and project information below identify whether a run uses the synthetic mechanics fixture or a documented public study.

```{r inputs}
counts <- read_count_matrix(config$counts_path)
metadata <- read_sample_metadata(config$metadata_path)
contrasts <- read_contrasts(config$contrasts_path)
metadata <- validate_analysis_inputs(counts, metadata, contrasts, config$design)

data.frame(
  project = config$project_title,
  data_status = config$data_label,
  genes = nrow(counts),
  samples = ncol(counts),
  design = config$design,
  contrasts = nrow(contrasts),
  stringsAsFactors = FALSE
)
knitr::kable(metadata, caption = "Sample metadata")
knitr::kable(contrasts, caption = "Declared comparisons")
```

# Filtering and model fit

Genes with total count below `r config$min_total_count` are excluded before fitting. DESeq2 estimates library-size factors and dispersions and fits the configured design.

```{r model}
dds <- prepare_deseq_dataset(counts, metadata, config$design, config$min_total_count)
dds <- fit_deseq_model(dds)
transformed <- transform_for_visualization(dds)

normalized_counts <- as.data.frame(DESeq2::counts(dds, normalized = TRUE))
normalized_counts$gene_id <- rownames(normalized_counts)
normalized_counts <- normalized_counts[, c("gene_id", setdiff(names(normalized_counts), "gene_id"))]
write.csv(normalized_counts, file.path(paths[["tables"]], "normalized_counts.csv"), row.names = FALSE)
write.csv(
  data.frame(
    project = config$project_title,
    data_status = config$data_label,
    design = config$design,
    fdr = config$fdr,
    min_total_count = config$min_total_count,
    stringsAsFactors = FALSE
  ),
  file.path(paths[["tables"]], "run_metadata.csv"),
  row.names = FALSE
)

qc_summary <- data.frame(
  sample_id = colnames(counts),
  library_size = colSums(counts),
  detected_genes = colSums(counts > 0),
  size_factor = DESeq2::sizeFactors(dds),
  row.names = NULL
)
knitr::kable(qc_summary, digits = 3, caption = "Library and normalization summary")
```

# Sample-level quality assessment

PCA uses the configured number of highest-variance variance-stabilized genes. The correlation heatmap uses all retained transformed genes.

```{r sample-qc, fig.width=7, fig.height=5}
pca_plot <- make_pca_plot(transformed, metadata, config$pca_top_genes, paste(config$data_label, "PCA", sep = ": "))
print(pca_plot)
save_ggplot(pca_plot, file.path(paths[["figures"]], "pca.png"))
correlation_path <- file.path(paths[["figures"]], "sample_correlation.png")
save_correlation_heatmap(transformed, correlation_path, paste(config$data_label, "sample correlation", sep = ": "))
knitr::include_graphics(normalizePath(correlation_path))
```

# Differential expression

Adjusted p-values use the Benjamini-Hochberg procedure. Direction indicates the sign of the numerator-versus-denominator log2 fold change among genes passing the configured FDR.

```{r differential-expression, results='asis', fig.width=7, fig.height=5}
contrast_results <- list()
for (index in seq_len(nrow(contrasts))) {
  comparison <- contrasts[index, , drop = FALSE]
  label <- comparison$label[[1L]]
  stem <- safe_file_stem(label)
  result <- run_deseq_contrast(dds, comparison, config$fdr)
  contrast_results[[label]] <- result
  write.csv(result, file.path(paths[["tables"]], paste0(stem, "_deseq2.csv")), row.names = FALSE)

  cat("\n\n## ", label, "\n\n", sep = "")
  summary_table <- as.data.frame(table(result$direction), stringsAsFactors = FALSE)
  names(summary_table) <- c("classification", "genes")
  print(knitr::kable(summary_table, caption = paste("FDR", config$fdr)))
  volcano <- make_volcano_plot(result, label, config$fdr, config$data_label)
  print(volcano)
  save_ggplot(volcano, file.path(paths[["figures"]], paste0(stem, "_volcano.png")))

  ma_plot <- make_ma_plot(result, label, config$fdr, config$data_label)
  print(ma_plot)
  save_ggplot(ma_plot, file.path(paths[["figures"]], paste0(stem, "_ma.png")))

  heatmap_path <- file.path(paths[["figures"]], paste0(stem, "_significant_genes_heatmap.png"))
  heatmap_summary <- save_significant_gene_heatmap(
    transformed, metadata, heatmap_path, config$heatmap_top_genes, result,
    config$fdr, config,
    paste(config$data_label, "significant genes across replicates", sep = ": ")
  )
  cat("\n\nThe heatmap shows ", length(heatmap_summary$selected_genes), " of ",
      heatmap_summary$available_significant, " genes passing adjusted p-value < ",
      config$fdr, ". Values are row-centered variance-stabilized counts; they are not z-scores.\n\n", sep = "")
  cat("\n\n![](figures/", basename(heatmap_path), ")\n\n", sep = "")
}
```

# Optional functional enrichment

Enrichment is disabled for the synthetic fixture. When enabled, the workflow maps the configured identifier type to Entrez IDs and analyzes significant up- and down-regulated genes separately against the set of tested genes. Failure to map identifiers produces empty outputs rather than a biological claim.

```{r enrichment}
if (isTRUE(config$enrichment$enabled)) {
  for (label in names(contrast_results)) {
    result <- contrast_results[[label]]
    stem <- safe_file_stem(label)
    for (direction in c("up", "down")) {
      selected <- result$gene_id[result$direction == direction]
      enrichment <- run_optional_enrichment(
        gene_ids = selected,
        universe_ids = result$gene_id[!is.na(result$pvalue)],
        direction = direction,
        contrast_label = label,
        config = config
      )
      write.csv(enrichment$go_bp, file.path(paths[["tables"]], paste0(stem, "_", direction, "_go_bp.csv")), row.names = FALSE)
      write.csv(enrichment$reactome, file.path(paths[["tables"]], paste0(stem, "_", direction, "_reactome.csv")), row.names = FALSE)
      write.csv(enrichment$mapping_coverage, file.path(paths[["tables"]], paste0(stem, "_", direction, "_mapping_coverage.csv")), row.names = FALSE)
    }
  }
} else {
  cat("Enrichment was not requested in this configuration.")
}
```

# Reproducibility

```{r session-info}
sessionInfo()
```
