---
title: "RNA-seq downstream analysis"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  config_path: "config/example_config.R"
  portfolio_snapshot: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5)
source(file.path("R", "helpers.R"))
resolve_project_path <- function(path) if (grepl("^/", path)) path else file.path(getwd(), path)
config <- load_analysis_config(params$config_path)
paths <- prepare_output_directories(resolve_project_path(config$output_dir))
warning_log <- character()
capture_warnings <- function(expr) withCallingHandlers(expr, warning = function(w) {
  warning_log <<- c(warning_log, conditionMessage(w)); invokeRestart("muffleWarning")
})

counts_path <- resolve_project_path(config$counts_path)
metadata_path <- resolve_project_path(config$metadata_path)
contrasts_path <- resolve_project_path(config$contrasts_path)
counts <- read_count_matrix(counts_path)
sample_metadata <- read_sample_metadata(metadata_path)
contrasts <- read_contrasts(contrasts_path)
sample_metadata <- validate_analysis_inputs(counts, sample_metadata, contrasts, config$design)

dds <- prepare_deseq_dataset(counts, sample_metadata, config$design, config$min_total_count)
dds <- capture_warnings(fit_deseq_model(dds))
transformed <- capture_warnings(transform_for_visualization(dds))

normalized_counts <- as.data.frame(DESeq2::counts(dds, normalized = TRUE), check.names = FALSE)
normalized_counts$gene_id <- rownames(normalized_counts)
normalized_counts <- normalized_counts[, c("gene_id", setdiff(names(normalized_counts), "gene_id"))]
write.csv(normalized_counts, file.path(paths[["tables"]], "normalized_counts.csv"), row.names = FALSE)

qc_summary <- data.frame(
  sample_id = colnames(counts), library_size = colSums(counts),
  detected_genes = colSums(counts > 0), size_factor = DESeq2::sizeFactors(dds),
  row.names = NULL
)
write.csv(qc_summary, file.path(paths[["tables"]], "sample_qc.csv"), row.names = FALSE)

contrast_results <- list()
summary_rows <- vector("list", nrow(contrasts))
for (index in seq_len(nrow(contrasts))) {
  comparison <- contrasts[index, , drop = FALSE]
  label <- comparison$label[[1L]]
  result <- capture_warnings(run_deseq_contrast(
    dds, comparison, config$fdr, input_counts = counts,
    min_total_count = config$min_total_count
  ))
  contrast_results[[label]] <- result
  summary_rows[[index]] <- summarize_contrast_result(result, label)
  write.csv(result, file.path(paths[["tables"]], paste0(safe_file_stem(label), "_deseq2.csv")),
            row.names = FALSE)
}
result_summary <- do.call(rbind, summary_rows)
write.csv(result_summary, file.path(paths[["tables"]], "contrast_summary.csv"), row.names = FALSE)

input_files <- c(counts = counts_path, metadata = metadata_path, contrasts = contrasts_path,
                 config = params$config_path)
write.csv(input_manifest(input_files), file.path(paths[["tables"]], "input_manifest.csv"), row.names = FALSE)
write.csv(software_versions(unique(c(required_core_packages(),
  if (isTRUE(config$gene_labels$enabled)) c("AnnotationDbi", config$gene_labels$organism_db) else character()))),
  file.path(paths[["tables"]], "software_versions.csv"), row.names = FALSE)
write.csv(code_manifest(c("R/helpers.R", "report/rnaseq_downstream.Rmd", "run_analysis.R")),
          file.path(paths[["tables"]], "code_manifest.csv"), row.names = FALSE)
writeLines(capture.output(dput(config)), file.path(paths[["tables"]], "resolved_config.txt"))
run_metadata <- data.frame(
  field = c("project", "data_status", "design", "fdr", "minimum_total_count",
            "input_genes", "retained_genes", "samples", "contrasts", "generated_utc"),
  value = c(config$project_title, config$data_label, config$design, config$fdr,
            config$min_total_count, nrow(counts), nrow(dds), ncol(counts), nrow(contrasts),
            format(Sys.time(), tz = "UTC", usetz = TRUE)), stringsAsFactors = FALSE
)
write.csv(run_metadata, file.path(paths[["tables"]], "run_metadata.csv"), row.names = FALSE)
```

# Analysis summary

**`r config$project_title`**  
**Data status:** `r config$data_label`

This report starts from a declared integer gene-count matrix; it does not perform read alignment or gene quantification. All comparisons come from the supplied contrast table.

```{r headline-summary}
knitr::kable(result_summary, digits = 4, caption = "Contrast-level result ledger summary")
```

`input_rows` is the full input feature universe. A missing adjusted p-value is never called non-significant: `all_zero`, `low_count_filtered`, `unavailable`, and `independent_filtered` remain separate from `tested_nonsignificant`. Benjamini-Hochberg adjustment and DESeq2 independent filtering are applied separately within each declared contrast.

# Inputs and design

```{r inputs}
knitr::kable(run_metadata, caption = "Resolved run metadata")
knitr::kable(sample_metadata, caption = "Sample metadata")
knitr::kable(contrasts, caption = "Declared comparisons")
knitr::kable(qc_summary, digits = 3, caption = "Library and normalization summary")
```

Genes with total count below `r config$min_total_count` are excluded from model fitting but retained in each result ledger. The design is `r config$design`; categorical metadata with numeric-only codes are rejected to avoid accidental continuous-variable semantics.

# Sample-level diagnostics

PCA uses the `r config$pca_top_genes` highest-variance variance-stabilized genes. The correlation matrix uses every retained transformed gene and prints Pearson coefficients in the figure and downloadable table.

```{r diagnostics}
pca_plot <- make_pca_plot(transformed, sample_metadata, config$pca_top_genes,
                          paste(config$data_label, "PCA", sep = ": "))
print(pca_plot)
save_ggplot(pca_plot, file.path(paths[["figures"]], "pca.png"))
correlation_path <- file.path(paths[["figures"]], "sample_correlation.png")
correlation <- save_correlation_heatmap(transformed, correlation_path,
                                        paste(config$data_label, "sample correlation", sep = ": "))
write.csv(correlation, file.path(paths[["tables"]], "sample_correlation.csv"))
knitr::include_graphics(normalizePath(correlation_path))
dispersion_path <- file.path(paths[["figures"]], "dispersion_estimates.png")
save_dispersion_plot(dds, dispersion_path, paste(config$data_label, "dispersion estimates", sep = ": "))
knitr::include_graphics(normalizePath(dispersion_path))
```

# Differential expression

Fold changes are unshrunk numerator-versus-denominator estimates. Heatmaps show a deterministic selection: lowest adjusted p-value, then largest absolute fold change, then gene ID; values are row-centered VST counts, not z-scores.

```{r differential-expression, results='asis'}
for (label in names(contrast_results)) {
  result <- contrast_results[[label]]
  stem <- safe_file_stem(label)
  cat("\n## ", label, "\n\n", sep = "")
  print(knitr::kable(result_summary[result_summary$contrast == label, ], digits = 4))
  ma_plot <- make_ma_plot(result, label, config$fdr, config$data_label)
  print(ma_plot)
  save_ggplot(ma_plot, file.path(paths[["figures"]], paste0(stem, "_ma.png")))
  volcano <- make_volcano_plot(result, label, config$fdr, config$data_label)
  print(volcano)
  save_ggplot(volcano, file.path(paths[["figures"]], paste0(stem, "_volcano.png")))
  heatmap_path <- file.path(paths[["figures"]], paste0(stem, "_significant_genes_heatmap.png"))
  heatmap_summary <- save_significant_gene_heatmap(
    transformed, sample_metadata, heatmap_path, config$heatmap_top_genes, result,
    config$fdr, config, paste(config$data_label, label, "significant genes", sep = ": ")
  )
  cat("\nThe heatmap shows ", length(heatmap_summary$selected_genes), " of ",
      heatmap_summary$available_significant, " significant genes.\n\n", sep = "")
  print(knitr::include_graphics(normalizePath(heatmap_path)))
}
```

# Optional functional enrichment

When enabled, up- and down-regulated genes are mapped separately to Entrez IDs. Mapping coverage is exported for every direction. Empty, completely unmapped, and partially mapped selections produce explicit statuses and schema-stable tables rather than biological claims.

```{r enrichment, results='asis'}
if (isTRUE(config$enrichment$enabled)) {
  for (label in names(contrast_results)) {
    result <- contrast_results[[label]]
    stem <- safe_file_stem(label)
    for (direction in c("up", "down")) {
      enrichment <- capture_warnings(run_optional_enrichment(
        result$gene_id[result$direction == direction],
        result$gene_id[result$retained_for_model & is.finite(result$pvalue)],
        direction, label, config
      ))
      write.csv(enrichment$go_bp, file.path(paths[["tables"]], paste0(stem, "_", direction, "_go_bp.csv")), row.names = FALSE)
      write.csv(enrichment$reactome, file.path(paths[["tables"]], paste0(stem, "_", direction, "_reactome.csv")), row.names = FALSE)
      write.csv(enrichment$mapping_coverage, file.path(paths[["tables"]], paste0(stem, "_", direction, "_mapping_coverage.csv")), row.names = FALSE)
      print(knitr::kable(enrichment$mapping_coverage))
    }
  }
} else cat("Enrichment was not requested in this configuration.\n")
```

# Downloads and reproducibility

```{r downloads, results='asis'}
warning_table <- data.frame(warning = unique(warning_log), stringsAsFactors = FALSE)
if (!nrow(warning_table)) warning_table <- data.frame(warning = "No warnings captured.")
write.csv(warning_table, file.path(paths[["tables"]], "warnings.csv"), row.names = FALSE)
session_text <- capture.output(sessionInfo())
writeLines(session_text, file.path(paths[["tables"]], "session_info.txt"))
if (isTRUE(params$portfolio_snapshot)) {
  cat("This saved portfolio snapshot links only to compact tracked results. Full tables are regenerated under the configured output directory.\n\n")
  snapshot_files <- c(
    "Contrast summary" = "../results/gse231397/contrast_summary.csv",
    "Published-target concordance" = "../results/gse231397/published_target_concordance.csv"
  )
  cat(paste0("- [", names(snapshot_files), "](", snapshot_files, ")", collapse = "\n"))
} else {
  table_files <- sort(list.files(paths[["tables"]]))
  cat(paste0("- [", table_files, "](tables/", table_files, ")", collapse = "\n"))
}
```

The manifests record resolved input paths and SHA-256 hashes, analysis-code hashes, resolved configuration, software versions, warnings, and the full R session. These are a transparent run record, not a claim that the environment is automatically restorable.

```{r session-info}
sessionInfo()
```
